Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-18 paper.tex

Figure~\ref{fig:breakpoint-distribution} shows the distribution of breakpoint
intervals and patterns of recurrent mutation along the genome in the Wide ARG
(see Figure~\ref{fig:long_arg_breakpoint_distribution} for the same information
of the Long ARG). We focus on the Wide ARG here as it covers roughly the same time
period as the analyses of \cite{Turakhia2022-it}, facilitating comparisons
of the results. To reduce the effect of artefactual recombinants, we
consider only the breakpoints associated with the 1,522 recombination nodes
that are ancestral to more than one sample. The mean length of these intervals
is 1,685 bases (median 962); the length distribution is summarised in the
inset histogram in Figure~\ref{fig:breakpoint-distribution}.

Figure~\ref{fig:breakpoint-distribution} [ TODO what can we really say here?
Other than the boundaries happen to lie on gene breaks, not an awful lot I
think?]

% This isn't really true any more, is it?
% There is a preponderance of breakpoint intervals spanning the 3’ end of the
% genome (Figure 4A), as noted previously by \cite{Turakhia2022-it}. Inspecting
% more closely, we find an enrichment in the number of intersecting intervals in
% the E gene, ORF8, ORF8-N intergenic region, and 3’ untranslated region (Figure
% 4B). Markedly, there is an abrupt drop in the number of intersecting intervals
% in the N gene following the peak in ORF8 towards the 3’ end.

Next, we investigate the potential recombination signals alongside the number
of mutations across the genome (Figure 4C). Regions with a higher density of
mutations contain more genetic variation, increasing the power to detect
recombination events. The position 28,271 (located in the ORF8-N intergenic
region) has a particularly high number of mutations in the samples (Figure 5).
At this position, mutations (n = 7,572), insertions (n = 5,782), and deletions
(n = 1,605) appear to occur frequently. One possible explanation is that this
homopolymeric region is prone to sequencing errors, giving rise to an
artifactual recombination signal. At the upstream position 27,972, numerous
mutations also occur. This position has been reported to be hypermutable
\citep{Jungreis2021-dh}.

We further consider the breakpoint intervals within the context of
recombination breakpoint sequence motifs. It is hypothesised that certain
palindromic breakpoint sequences (specifically, CAGAC and CAGAT) promote
template switching during replication in SARS-CoV-2 via formation of
base-paired stem loops in the genome structure \citep{Gallaher2020-lb}. There
are 87 occurrences of these two breakpoint sequences in the Wuhan-Hu-1/2019
reference sequence (Supplementary Table 5). We observe that XX (XX\%) of the
breakpoint intervals of the HMM-consistent recombinants span at least one of
the breakpoint sequences.

>>>>

Figure~\ref{fig:breakpoint-distribution} shows the distribution of breakpoint
intervals and patterns of recurrent mutation along the genome in the Wide ARG
(see Figure~\ref{fig:long_arg_breakpoint_distribution} for the same information
of the Long ARG). We focus on the Wide ARG here as it covers roughly the same time
period as the analyses of \cite{Turakhia2022-it}, facilitating comparisons
of the results. To reduce the effect of artefactual recombinants, we
consider only the breakpoints associated with the 1,522 recombination nodes
that are ancestral to more than one sample. The mean length of these intervals
is 1,685 bases (median 962), and the length distribution is summarised in the
inset histogram in Figure~\ref{fig:breakpoint-distribution}.

While the detailed interpretation of Figure~\ref{fig:breakpoint-distribution}
requires further study, a few key properties stand out.[TODO, write this
better. ] Firstly, in several cases there is a clear relationship between
gene boundaries and the breakpoint intervals. For example, ORF7a and OR7b
show sharp differences in the number of intersecting intervals within
these genes. This could be due to a number of factors, such as [BIOLOGY]
or the presence of sites with artefactually high numbers of mutations
in the intergenic regions (see the next paragraph for more discussion).
Secondly, we see substantially fewer intervals intersecting with the N
gene than other regions, in agreement with the results of
\cite{Turakhia2022-it}.

It is important to note that the exact endpoints of intervals can be
somewhat arbitrary, since they are defined by exact comparisons
of a recombinant's parents. The presence of problematic sites that are enriched
for errors will therefore tend to artificially truncate breakpoint intervals
at these sites.
It is therefore helpful to compare the number of interecting intervals
with the number of mutations per site in Figure~\ref{fig:breakpoint-distribution}.
For example, position 28,271 (located in the ORF8-N intergenic
region) has the largest number of mutatations and corresponds to the
left endpoint of 9 intervals and the right of 57.
The 7,572 mutations  (including 5,782 insertions and 1,605 deletions)
at this site are a likely indicator that this
homopolymeric region is prone to sequencing errors
and should be included in the list of ``problematic sites'' that
are excluded from analysis (Section~\ref{sec:data_preprocessing}).
% There can be good biological reasons for this too though
However, hypermutable sites such as 27,972 \citep{Jungreis2021-dh}
have the same effect of truncating intervals. At this site we have
770 mutations (including 425 C$>$T, 314 T$>$C) with potential signals
of polymorphism maintained by selection [FIXME!!], and 58 intervals
with endpoints within one base of this site.

% JK Dropping this for now:
% We further consider the breakpoint intervals within the context of
% recombination breakpoint sequence motifs. It is hypothesised that certain
% palindromic breakpoint sequences (specifically, CAGAC and CAGAT) promote
% template switching during replication in SARS-CoV-2 via formation of
% base-paired stem loops in the genome structure \citep{Gallaher2020-lb}. There
% are 87 occurrences of these two breakpoint sequences in the Wuhan-Hu-1/2019
% reference sequence (Supplementary Table 5). We observe that XX (XX\%) of the
% breakpoint intervals of the HMM-consistent recombinants span at least one of
% the breakpoint sequences.
